Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-08 dev/notebooks/gravitational_generator.qmd
2/3 real world data done

```{julia}
x = select_factual(counterfactual_data, rand(1:size(X)[2])) 
y = round(probs(M, x)[1])
target = ifelse(y==1.0,0.0,1.0) # opposite label as target
```

```{julia}
Λ₂ = [0.1, 1, 5]
counterfactuals = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = GravitationalGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals = vcat(
      counterfactuals...,
      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict)
    )
end
```

```{julia}
theme(:wong)
T = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals[i]  
    plt = plot(counterfactual, plot_up_to=T, title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1200,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator_strict.png"))
```

@@

```{julia}
theme(:wong)
T = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals[i]  
    plt = plot(counterfactual, plot_up_to=T, title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1400,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator.png"))
```

@@

>>>>

```{julia}
x = select_factual(counterfactual_data, rand(1:size(X)[2])) 
y = round(probs(M, x)[1])
target = ifelse(y==1.0,0.0,1.0) # opposite label as target
T = 100
```

```{julia}
Λ₂ = [0.1, 1, 5]
counterfactuals = []
generators = []
for λ₂ ∈ Λ₂  
    λ = [0.1, λ₂]
    generator = GravitationalGenerator(λ=λ)
    generators = vcat(generators..., generator)
    counterfactuals = vcat(
      counterfactuals...,
      generate_counterfactual(x, target, counterfactual_data, M, generator; convergence=:strict, T=T)
    )
end
```

```{julia}
theme(:wong)
T_ = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals[i]  
    plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1200,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator_strict.png"))
```

@@

```{julia}
theme(:wong)
T_ = 500
plts = []
for i ∈ 1:length(Λ₂)
    λ₂ = Λ₂[i]
    counterfactual = counterfactuals[i]  
    plt = plot(counterfactual, plot_up_to=minimum([T,T_]), title="λ₂=$(λ₂)")
    x1,x2 = generators[i].centroid[1], generators[i].centroid[2]
    scatter!(plt, [x1], [x2], colour=:purple, label="attractor")
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(1400,300), layout=(1,3))
savefig(plt, joinpath(www_path,"gravitational_generator.png"))
```

@@

```{julia}

```

### 2022-08-08 dev/notebooks/real_world.qmd
2/3 real world data done

```{julia}
catalogue = AlgorithmicRecourseDynamics.Data.load_real_world()
choices = [:give_me_some_credit_balanced, :housing]
data_sets = filter(p -> p[1] in choices, catalogue)
```

```{julia}
using CounterfactualExplanations.DataPreprocessing: unpack
bs = 50
function data_loader(data::CounterfactualData)
    X, y = unpack(data)
    data = Flux.DataLoader((X,y),batchsize=bs)
    return data
end
model_params = (batch_norm=false,n_hidden=128,dropout=true,p_dropout=0.1)
```

```{julia}
models = [:FluxModel, :FluxEnsemble]
generators = Dict(
    :Greedy=>GreedyGenerator(), 
    :Generic=>GenericGenerator(),
    :REVISE=>REVISEGenerator(),
    :DICE=>DiCEGenerator(),
    :Gravitational=>GravitationalGenerator(),
)
```

```{julia}
experiments = set_up_experiments(data_sets,models,generators; pre_train_models=100, model_params=model_params, data_loader=data_loader)
```

@@

@@

```{julia}
n_evals = 1
n_rounds = 10
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 2
n_bootstrap = 1
using Serialization
results = run_experiments(
    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap
)
Serialization.serialize(joinpath(output_path,"results.jls"),results)
```

@@

```{julia}
n_evals = 1
n_rounds = 10
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 2
n_bootstrap = 1
using Serialization
results = run_experiments(
    experiments;
    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
    convergence=:strict, save_name_suffix="strict"
)
Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
```

>>>>

```{julia}
max_obs = 10000
data_sets = AlgorithmicRecourseDynamics.Data.load_real_world(max_obs)
```

```{julia}
using CounterfactualExplanations.DataPreprocessing: unpack
bs = 50
function data_loader(data::CounterfactualData)
    X, y = unpack(data)
    data = Flux.DataLoader((X,y),batchsize=bs)
    return data
end
model_params = (batch_norm=false,n_hidden=32,n_layers=3,dropout=true,p_dropout=0.25)
```

```{julia}
models = [
    :LogisticRegression, 
    :FluxModel, 
    :FluxEnsemble
]
generators = Dict(
    :Greedy=>GreedyGenerator(), 
    :Generic=>GenericGenerator(),
    :REVISE=>REVISEGenerator(),
    :DICE=>DiCEGenerator(),
    :Gravitational=>GravitationalGenerator(λ=[0.1,2]),
)
```

```{julia}
experiments = set_up_experiments(
    data_sets,models,generators; 
    pre_train_models=100, model_params=model_params, 
    data_loader=data_loader
)
```

@@

```{julia}
n_evals = 5
n_rounds = 50
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 5
n_bootstrap = 1
T = 500
using Serialization
results = run_experiments(
    experiments;
    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap, T=T,
    convergence=:strict, save_name_suffix="strict"
)
Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
```

@@

```{julia}
n_evals = 5
n_rounds = 50
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 5
n_bootstrap = 1
using Serialization
results = run_experiments(
    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap
)
Serialization.serialize(joinpath(output_path,"results.jls"),results)
```

@@

### Plots

```{julia}
# plot_dict = Dict()
# for (data_name, res) in results
#     output_data = res.output
#     metrics = unique(output_data.name)
#     plots = Dict(metric => plot(res, metric) for metric in metrics)
#     plot_dict[data_name] = plots
# end
```

### 2022-08-08 dev/notebooks/synthetic.qmd
2/3 real world data done

```{julia}
n_evals = 1
n_rounds = 1
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 1
n_bootstrap = 1
using Serialization
results = run_experiments(
    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds,n_folds=n_folds,n_bootstrap=n_bootstrap
)
Serialization.serialize(joinpath(output_path,"results.jls"),results)
```

@@

@@

```{julia}
n_evals = 1
n_rounds = 1
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 1
n_bootstrap = 1
using Serialization
results = run_experiments(
    experiments;
    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
    convergence=:strict, save_name_suffix="strict"
)
Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
```

@@

>>>>

```{julia}
n_evals = 10
n_rounds = 50
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 5
n_bootstrap = 1
using Serialization
results = run_experiments(
    experiments;save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds,n_folds=n_folds,n_bootstrap=n_bootstrap
)
Serialization.serialize(joinpath(output_path,"results.jls"),results)
```

@@

```{julia}
plot_dict = Dict()
```

@@

```{julia}
n_evals = 10
n_rounds = 50
evaluate_every = Int(round(n_rounds/n_evals))
n_folds = 5
n_bootstrap = 1
using Serialization
results = run_experiments(
    experiments;
    save_path=output_path,evaluate_every=evaluate_every,n_rounds=n_rounds, n_folds=n_folds, n_bootstrap=n_bootstrap,
    convergence=:strict, save_name_suffix="strict"
)
Serialization.serialize(joinpath(output_path,"results_strict.jls"),results)
```

@@

```{julia}
plot_dict = Dict()
for (data_name, res) in results
    output_data = res.output
    metrics = unique(output_data.name)
    plots = Dict(metric => plot(res, metric) for metric in metrics)
    plot_dict[data_name] = plots
end
```

### 2022-08-08 dev/paper/paper.tex
2/3 real world data done

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Endogenous Dynamics in Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Endogenous Dynamics in Algorithmic Recourse}
\author{\author{
\IEEEauthorblockN{Patrick Altmeyer}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
p.altmeyer[at]tudelft.nl}
\and
\IEEEauthorblockN{Giovan Angela}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
g.j.a.angela[at]student.tudelft.nl}
\and
\IEEEauthorblockN{Aleksander Buszydlik}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
a.j.buszydlik[at]student.tudelft.nl}
\and
\IEEEauthorblockN{Karol Dobiczek}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
k.t.dobiczek[at]student.tudelft.nl}
\and
\IEEEauthorblockN{Cynthia C. S. Liem}
\IEEEauthorblockA{\textit{EEMCS} \\
\textit{Delft University of Technology}\\
Delft, Netherlands \\
c.c.s.liem[at]tudelft.nl}
}}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, boxrule=0pt, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

The \textbf{Operating Point} or \textbf{elbow} of the \(k\)-means
algorithm is used to measure whether the counterfactual instances tend
to lie close to the positively labelled training samples or make up a
new, separate cluster of data points within the positive class. In the
latter case the operating point of the \(k\)-means algorithm can be
expected to change. There exist several algorithms for the automated
discovery of elbow points and we resort to the Kneedle method
\protect\hyperlink{ref-satopaa2011finding}{{[}26{]}}. Kneedle selects
the optimal value of \(k\) by rotating the inertia curve onto the
horizontal axis and selecting a local minimum. While working with the
synthetic datasets we also use domain knowledge to confirm the results
of Kneedle through the visual inspection of data.

@@

As an alternative to PP MMD we use a pseudo-distance for the
\textbf{Disagreement Coefficient} (Disagreement). This metric was
introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}27{]}} and
estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
classifiers do not agree on the predicted outcome for a randomly chosen
sample. Thus, it is not relevant whether the classification is correct
according to the ground truth, but only whether the sample lies on the
same side of the two respective decision boundaries. In our context,
this metric quantifies the overlap between the initial model (trained
before the application of recourse) and the updated model. A
Disagreement Coefficient unequal to zero is indicative of a model shift.
The opposite is not true: even if the Disagreement Coefficient is equal
to zero a model shift may still have occured. This is one reason for why
PP MMD is our our preferred metric.

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-satopaa2011finding}{}}%
\CSLLeftMargin{{[}26{]} }%
\CSLRightInline{V. Satopaa, J. Albrecht, D. Irwin, and B. Raghavan,
{``Finding a" kneedle" in a haystack: Detecting knee points in system
behavior,''} in \emph{2011 31st international conference on distributed
computing systems workshops}, 2011, pp. 166--171.}

\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
\CSLLeftMargin{{[}27{]} }%
\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
agnostic active learning,''} in \emph{Proceedings of the 24th
international conference on machine learning}, 2007, pp. 353--360.}

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithm}% http://ctan.org/pkg/algorithms
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{placeins}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Endogenous Macrodynamics in Algorithmic Recourse},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Endogenous Macrodynamics in Algorithmic Recourse}
\author{}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden, sharp corners, breakable]}{\end{tcolorbox}}\fi

@@

@@

As an alternative to PP MMD we use a pseudo-distance for the
\textbf{Disagreement Coefficient} (Disagreement). This metric was
introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}26{]}} and
estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two
classifiers do not agree on the predicted outcome for a randomly chosen
sample. Thus, it is not relevant whether the classification is correct
according to the ground truth, but only whether the sample lies on the
same side of the two respective decision boundaries. In our context,
this metric quantifies the overlap between the initial model (trained
before the application of recourse) and the updated model. A
Disagreement Coefficient unequal to zero is indicative of a model shift.
The opposite is not true: even if the Disagreement Coefficient is equal
to zero a model shift may still have occured. This is one reason for why
PP MMD is our our preferred metric.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Shift of focus from individual to group of individuals (related:
  https://www.researchgate.net/publication/353073138\_Generating\_Collective\_Counterfactual\_Explanations\_in\_Score-Based\_Classification\_via\_Mathematical\_Optimization)
\item
  Convergence criterium matters: terminating once threshold probability
  is reached may not be optimal (see e.g.~REVISE)
\item
  Optimizer choice matters: dimensionality is typically low, so no
  obvious benefit to using ADAM.

  \begin{itemize}
  \tightlist
  \item
    This might be better placed in JuliaCon proceedings, perhaps backed
    by small blog post on the matter.
  \end{itemize}
\item
  Mitigating strategy: penaliye distance from centroid.
\end{enumerate}

@@

\hypertarget{data}{%
\subsubsection{Data}\label{data}}

The number of real-world datasets considered in this work is limited,
constrained by the fact that at the time of writing
\texttt{CounterfactualExplanations.jl} only supported continuous
features.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hanneke2007bound}{}}%
\CSLLeftMargin{{[}26{]} }%
\CSLRightInline{S. Hanneke, {``A bound on the label complexity of
agnostic active learning,''} in \emph{Proceedings of the 24th
international conference on machine learning}, 2007, pp. 353--360.}
